Lead distance under a pickoff limit in Major League Baseball:\\A sequential game model

### 2025-12-24 articles/arxiv/paper.tex
fix typos

    For background, the stolen base is an exciting play in which a runner takes off on the pitch, attempting to reach the next base before the catcher can throw the to a fielder at the base to tag the runner out \parencite{mlb_baseball_2025}. Whether attempting a stolen base or not, the runner takes a {\it lead} from their base, commonly positioning themselves 8--12 feet away from their base, toward the next base. The longer the lead, the shorter the distance needed to run to reach the next base. The pitcher's primary deterrent against long leads is a {\it pickoff attempt}, in which they throw the ball to a fielder at the runner's base, attempting a tag before the runner returns safely to the base.

@@

    Section \ref{sec:sg-model} uses a two-payer zero-sum sequential game to model the evolving interaction between the runner and the pitcher.  Our model captures the vital aspects of the play while retaining analytical and computational tractability. Our model is closely related to the class of stochastic/dynamic Stackelberg games \parencite{goktas_zero-sum_2022,li_review_2017,lopez_stationary_2023,vasal_sequential_2022,vorobeychik_computing_2012}, which combines the elements of Stackelberg and stochastic games. The main difference between our infinite-horizon model and the stochastic Stackelberg games literature is that we do not discount the rewards. The rationale for this modeling choice is that our optimization objective---expected runs in the inning---by definition includes no discount. We establish a maximin Bellman equation for our sequential game and propose a variant of value iteration and policy iteration to determine the optimal strategies for runner and pitcher.

@@

    The goal of the batting team is to select a policy that will maximize the expected total runs scored by them. At the same time, the fielding team minimizes the expected total runs scored by the batting team by employing a suitable policy. Taking the rational behavior of both the players into account, the maximum runs scored by the batting team when a play begins from state $s \in {\cal S}$ is given as 
   \begin{equation}
        \label{eqn:sg_val}
        V^*(s) \defeq \max_{\pi_\text{R} \in \Pi_\text{R}}\min_{\pi_\text{P} \in \Pi_\text{P}}V^{\pi_\text{R},\pi_\text{P}}(s) \ \forall s \in S.
   \end{equation}
   We call $V^*$ the \emph{optimal value function}. In our context, $V^*$ is the maximum expected total runs scored in an inning by the batting team, which, by the zero-sum property of the game, equals the minimum expected total runs incurred by the fielding team. A policy pair $(\pi_\text{R}^*,\pi_\text{P}^*) \in \Pi_\text{R} \times \Pi_\text{P}$ such that $V^{\pi_\text{R}^*,\pi_\text{P}^*}(s) = V^*(s) \ \forall s \in {\cal S}$ is called an equilibrium policy. That is, $\pi_\text{R}^*$ is a policy for the batting team that maximizes the expected total runs scored by them, and $\pi_\text{P}^*$ is a policy for the fielding team that minimizes the expected total runs incurred by them. In other words, $\pi_\text{R}^*$ is a optimal policy for the batting team and $\pi_\text{P}^*$ is an optimal policy for the fielding team.

@@

    When the play begins, we use $\ell_i \in \mathbb{R}^+$ to denote the lead distance (in feet) taken by the first-base runner; and we use $p_i \in \{0, 1\}$ to denote whether the pitcher attempts (1) or does not attempt (0) a pickoff. We use $r_i \in \{\mbox{P}^+,\, \mbox{P}^-,\, \mbox{S}^+,\, \mbox{S}^-,\, \mbox{N}\} \equiv \mathcal{R}$ to denote the runner outcome, where
    \begin{align}
    \label{eqn:runner-outcome}
      \begin{split}
        & \mbox{P}^+  \mbox{ represents a successful pickoff attempt (runner is out);}\\
        & \mbox{P}^-  \mbox{ represents an unsuccessful pickoff attempt (runner is safe);}\\
        & \mbox{S}^+  \mbox{ represents a successful stolen base attempt (runner is safe);}\\
        & \mbox{S}^-  \mbox{ represents an usuccessful stolen base attempt (runner is out); and}\\
        & \mbox{N}~   \mbox{ represents no runner action (a pitch with no stolen base attempt).}
      \end{split}
    \end{align}
    If there is no runner on first base, then $\ell_i$ and $r_i$ are undefined.

@@

        We model $\phi_i$ using a mixed-effects logistic regression with fixed effects for balls, strikes, outs, disengagements (as a categorical), and lead distance; and with a random effect for the pitcher.
        \begin{align}
          \label{eqn:prob-po-attempt}
          \begin{split}
            \log\left(\frac{\phi_i}{1 - \phi_i}\right) &= \alpha + \beta^B c_i^B + \beta^S c_i^S + \beta^O o_i + \beta^D_{d_i} + \beta^L \ell_i + \gamma^P_{h_i^P}\\
            \gamma^P_{h} &\sim \mathcal{N}(0, \sigma^2_P) \hspace{4mm} \mbox{\it i.i.d.} \hspace{4mm} \forall h \in \mathcal{H}^P.
          \end{split}
        \end{align}
        This model has eight fixed, unknown parameters: the intercept $\alpha$; the five slopes $\beta^B$, $\beta^S$, $\beta^O$, $\beta^D$, and $\beta^L$; and the two variance parameters $\sigma^2_R$ and $\sigma^2_P$. The number of random effects is $|\mathcal{H}^R| + |\mathcal{H}^P|$.

@@

      \begin{table}
        \small
        \center
        \input{../../output/tables/model_summary.tex}
        \caption{\it Summary of estimated runner outcome probability models. We estimated four different generalized (binomial) linear mixed-effects models for pickoff (PO) attempt and success and for stolen base (SB) attempt and success. For fixed effects we report the estimated effect with one standard error, and for random effects we report the estimated standard deviation term. For 2022, we encoded disengagements as always zero to reflect no pickoff limit.}
        \label{tab:model-summary}
      \end{table}

@@

      \begin{figure}[H]
        \centering
        \includegraphics[width = 0.4\textwidth]{effect_catcher_light.pdf}
        \includegraphics[width = 0.4\textwidth]{effect_runner_light.pdf}
        \caption{
          \it
          (Left) The relationship between arm strength and catcher combined effect on stolen base success probability (relative to average, on the log-odds scale). The combined effect is the sum of the fixed effect due to arm strength and the catcher random effect.
          (Right) The relationship between sprint speed and runner combined effect on stolen base success probability (relative to average, on the log-odds scale). The combined effect is the sum of the fixed effect due to sprint speed and the runner random effect.
        }
        \label{fig:random-effect}
      \end{figure}

@@

      \begin{table}[H]
        \centering
        \input{../../output/tables/lead_by_count_two_player.tex}
        \caption{
          \it The optimal lead distance (in feet) by count and prior disengagements under the two-player sequential game, assuming an average runner on first base facing an average battery (pitcher and catcher) with no outs and third base empty.
        }
        \label{tab:lead-by-count-two-player}
      \end{table}

@@

      \begin{table}[H]
        \centering
        \input{../../output/tables/lead_by_outs_two_player.tex}
        \caption{
          \it The optimal lead distance (in feet) by outs and prior disengagements under the two-player sequential game, assuming an average runner on first base facing an average battery (pitcher and catcher) with an 0-0 count and third base empty.
        }
        \label{tab:lead-by-outs-two-player}
      \end{table}

@@

      \begin{table}[H]
        \centering
        \input{../../output/tables/lead_by_count_one_player.tex}
        \caption{
          \it The optimal lead distance (in feet) by count and prior disengagements under the single-player MDP, assuming an average runner on first base facing an average battery (pitcher and catcher) with no outs and third base empty.
        }
        \label{tab:lead-by-count-one-player}
      \end{table}

@@

      \begin{table}[H]
        \centering
        \input{../../output/tables/lead_by_outs_one_player.tex}
        \caption{
          \it The optimal lead distance (in feet) by outs and prior disengagements under the single-player MDP, assuming an average runner on first base facing an average battery (pitcher and catcher) with an 0-0 count and third base empty.
        }
        \label{tab:lead-by-outs-one-player}
      \end{table}

@@

      \begin{table}[H]
        \centering
        \input{../../output/tables/lead_by_players.tex}
        \caption{
          \it Optimal lead distance (in feet) by battery/runner skill and prior disengagements, for an 0-0 count with no outs and third base empty. For example, the hypothetical 90$^{th}$-percentile battery represents the 90$^{th}$ percentile at attempting pickoffs, successfully executing pickoffs, suppressing stolen base attempts, and catching would-be base stealers.
        }
        \label{tab:lead-by-players}
      \end{table}

>>>>

    For background, the stolen base is an exciting play in which a runner takes off on the pitch, attempting to reach the next base before the catcher can throw the ball to a fielder at the base to tag the runner out \parencite{mlb_baseball_2025}. Whether attempting a stolen base or not, the runner takes a {\it lead} from their base, commonly positioning themselves 8--12 feet away from their base, toward the next base. The longer the lead, the shorter the distance needed to run to reach the next base. The pitcher's primary deterrent against long leads is a {\it pickoff attempt}, in which they throw the ball to a fielder at the runner's base, attempting a tag before the runner returns safely to the base.

@@

    Section \ref{sec:sg-model} uses a two-player zero-sum sequential game to model the evolving interaction between the runner and the pitcher.  Our model captures the vital aspects of the play while retaining analytical and computational tractability. Our model is closely related to the class of stochastic/dynamic Stackelberg games \parencite{goktas_zero-sum_2022,li_review_2017,lopez_stationary_2023,vasal_sequential_2022,vorobeychik_computing_2012}, which combines the elements of Stackelberg and stochastic games. The main difference between our infinite-horizon model and the stochastic Stackelberg games literature is that we do not discount the rewards. The rationale for this modeling choice is that our optimization objective---expected runs in the inning---by definition includes no discount. We establish a maximin Bellman equation for our sequential game and propose a variant of value iteration and policy iteration to determine the optimal strategies for runner and pitcher.

@@

    The goal of the batting team is to select a policy that will maximize the expected total runs scored by them. At the same time, the fielding team minimizes the expected total runs scored by the batting team by employing a suitable policy. Taking the rational behavior of both the players into account, the maximum runs scored by the batting team when a play begins from state $s \in {\cal S}$ is given as 
   \begin{equation}
        \label{eqn:sg_val}
        V^*(s) \defeq \max_{\pi_\text{R} \in \Pi_\text{R}}\min_{\pi_\text{P} \in \Pi_\text{P}}V^{\pi_\text{R},\pi_\text{P}}(s) \ \forall s \in S.
   \end{equation}
   We call $V^*$ the \emph{optimal value function}. In our context, $V^*$ is the maximum expected total runs scored in an inning by the batting team, which, by the zero-sum property of the game, equals the minimum expected total runs incurred by the fielding team. A policy pair $(\pi_\text{R}^*,\pi_\text{P}^*) \in \Pi_\text{R} \times \Pi_\text{P}$ such that $V^{\pi_\text{R}^*,\pi_\text{P}^*}(s) = V^*(s) \ \forall s \in {\cal S}$ is called an equilibrium policy. That is, $\pi_\text{R}^*$ is a policy for the batting team that maximizes the expected total runs scored by them, and $\pi_\text{P}^*$ is a policy for the fielding team that minimizes the expected total runs incurred by them. In other words, $\pi_\text{R}^*$ is an optimal policy for the batting team and $\pi_\text{P}^*$ is an optimal policy for the fielding team.

@@

    When the play begins, we use $\ell_i \in \mathbb{R}^+$ to denote the lead distance (in feet) taken by the first-base runner; and we use $p_i \in \{0, 1\}$ to denote whether the pitcher attempts (1) or does not attempt (0) a pickoff. We use $r_i \in \{\mbox{P}^+,\, \mbox{P}^-,\, \mbox{S}^+,\, \mbox{S}^-,\, \mbox{N}\} \equiv \mathcal{R}$ to denote the runner outcome, where
    \begin{align}
    \label{eqn:runner-outcome}
      \begin{split}
        & \mbox{P}^+  \mbox{ represents a successful pickoff attempt (runner is out);}\\
        & \mbox{P}^-  \mbox{ represents an unsuccessful pickoff attempt (runner is safe);}\\
        & \mbox{S}^+  \mbox{ represents a successful stolen base attempt (runner is safe);}\\
        & \mbox{S}^-  \mbox{ represents an unsuccessful stolen base attempt (runner is out); and}\\
        & \mbox{N}~   \mbox{ represents no runner action (a pitch with no stolen base attempt).}
      \end{split}
    \end{align}
    If there is no runner on first base, then $\ell_i$ and $r_i$ are undefined.

@@

        We model $\phi_i$ using a mixed-effects logistic regression with fixed effects for balls, strikes, outs, disengagements (as a categorical), and lead distance; and with a random effect for the pitcher.
        \begin{align}
          \label{eqn:prob-po-attempt}
          \begin{split}
            \log\left(\frac{\phi_i}{1 - \phi_i}\right) &= \alpha + \beta^B c_i^B + \beta^S c_i^S + \beta^O o_i + \beta^D_{d_i} + \beta^L \ell_i + \gamma^P_{h_i^P}\\
            \gamma^P_{h} &\sim \mathcal{N}(0, \sigma^2_P) \hspace{4mm} \mbox{\it i.i.d.} \hspace{4mm} \forall h \in \mathcal{H}^P.
          \end{split}
        \end{align}
        This model has seven fixed, unknown parameters: the intercept $\alpha$; the five slopes $\beta^B$, $\beta^S$, $\beta^O$, $\beta^D$, and $\beta^L$; and the variance parameter $\sigma^2_P$. The number of random effects is $|\mathcal{H}^P|$.

@@

      \begin{table}
        \footnotesize
        \center
        \input{../../output/tables/model_summary.tex}
        \caption{\it Summary of estimated runner outcome probability models. We estimated four different generalized (binomial) linear mixed-effects models for pickoff (PO) attempt and success and for stolen base (SB) attempt and success. For fixed effects we report the estimated effect with one standard error, and for random effects we report the estimated standard deviation term. For 2022, we encoded disengagements as always zero to reflect no pickoff limit.}
        \label{tab:model-summary}
      \end{table}

@@

      \begin{figure}
        \centering
        \includegraphics[width = 0.4\textwidth]{effect_catcher_light.pdf}
        \includegraphics[width = 0.4\textwidth]{effect_runner_light.pdf}
        \caption{
          \it
          (Left) The relationship between arm strength and catcher combined effect on stolen base success probability (relative to average, on the log-odds scale). The combined effect is the sum of the fixed effect due to arm strength and the catcher random effect.
          (Right) The relationship between sprint speed and runner combined effect on stolen base success probability (relative to average, on the log-odds scale). The combined effect is the sum of the fixed effect due to sprint speed and the runner random effect.
        }
        \label{fig:random-effect}
      \end{figure}

@@

      \begin{table}
        \centering
        \input{../../output/tables/lead_by_count_two_player.tex}
        \caption{
          \it The optimal lead distance (in feet) by count and prior disengagements under the two-player sequential game, assuming an average runner on first base facing an average battery (pitcher and catcher) with no outs and third base empty.
        }
        \label{tab:lead-by-count-two-player}
      \end{table}

@@

      \begin{table}
        \centering
        \input{../../output/tables/lead_by_outs_two_player.tex}
        \caption{
          \it The optimal lead distance (in feet) by outs and prior disengagements under the two-player sequential game, assuming an average runner on first base facing an average battery (pitcher and catcher) with an 0-0 count and third base empty.
        }
        \label{tab:lead-by-outs-two-player}
      \end{table}

@@

      \begin{table}[b]
        \centering
        \input{../../output/tables/lead_by_count_one_player.tex}
        \caption{
          \it The optimal lead distance (in feet) by count and prior disengagements under the single-player MDP, assuming an average runner on first base facing an average battery (pitcher and catcher) with no outs and third base empty.
        }
        \label{tab:lead-by-count-one-player}
      \end{table}

@@

      \begin{table}
        \centering
        \input{../../output/tables/lead_by_outs_one_player.tex}
        \caption{
          \it The optimal lead distance (in feet) by outs and prior disengagements under the single-player MDP, assuming an average runner on first base facing an average battery (pitcher and catcher) with an 0-0 count and third base empty.
        }
        \label{tab:lead-by-outs-one-player}
      \end{table}

@@

      \begin{table}
        \centering
        \input{../../output/tables/lead_by_players.tex}
        \caption{
          \it Optimal lead distance (in feet) by battery/runner skill and prior disengagements, for an 0-0 count with no outs and third base empty. For example, the hypothetical 90$^{th}$-percentile battery represents the 90$^{th}$ percentile at attempting pickoffs, successfully executing pickoffs, suppressing stolen base attempts, and catching would-be base stealers.
        }
        \label{tab:lead-by-players}
      \end{table}
